---
title: "Pathways in MS1"
author: "Clara Jégousse"
date: "`r Sys.Date()`"
output: html_document
---


# Objectives

According to Neftaly, the objectives are:

To retrieve the phosphorylation values (scaled abundance) of phosphopeptides found for all of these proteins in the MS1 dataset at all time points.

To plot small heatmaps for each pathway with all the phosphopeptides found for their proteins in the strains Guy11 and Dpmk1

# Libraries

```{r, warning=FALSE, message=FALSE}
library(here)
library(readr)
library(dplyr)
```


# Load the data

## Selected proteins and pathways

Let's start by loading the selected phosphopeptides:

```{r, warning=FALSE}
pathways_filepath = paste0(here::here(), "/cleaned/013_pathways.xlsx")
pathways = readxl::read_xlsx(pathways_filepath)
```

There are 91 proteins representing 6 pathways

```{r}
dim(pathways)
unique(pathways$pathway) %>% length()
unique(pathways$id_number) %>% length()
```


## MS1 data


### Load and clean the data

```{r}
ms1_filepath = paste0(here::here(), "/cleaned/ms1_cleaned.csv")
cleaned = readr::read_csv(ms1_filepath)
```

### Build the matrix

The long dataframe needs pivot-ing into a matrix. Scaling is applied too. 

```{r}
ms1_tb <- cleaned %>%
  mutate(sample_time = paste(sample, time, sep="-"),
         id = paste(`Positions in Master Proteins`, `Modifications`, sep="-")) %>%
  select(id, sample_time, value) %>%
  pivot_wider(names_from = "sample_time", values_from = "value" )

m = ms1_tb
rows <- m$id
m$id <- NULL
raw_m <- as.matrix(m)
```

### Replace missing values

We can also replace NA with the minimum observed value on the whole abundance data - before scaling and distance measuring.

```{r}
m <- raw_m
m[is.na(m)] <- min(m, na.rm=TRUE)
```

### Scaling

```{r}
m <- t(apply(m, 1, function(x)(x-min(x, na.rm=TRUE))/(max(x, na.rm=TRUE)-min(x,na.rm=TRUE))))
```


# Filtering 

```{r}
colnames(m)
rownames(m) = ms1_tb$id
```


```{r}
dim(m)
```

# HEATMAPS

## CpkA

```{r}
selected_pathway = pathways %>% filter(pathway == 'CpkA')

selected_ms1 = m[grepl(paste(selected_pathway$id_number, collapse="|"), ms1_tb$id),]
row_infos = data.frame(m_rownames = rownames(selected_ms1))
row_infos = row_infos %>% mutate(
  id_number = stringr::str_sub(m_rownames, end = 9)
  )
row_infos = left_join(row_infos, selected_pathway, by = "id_number", keep = FALSE)

mdata = list(row_info = row_infos, data = selected_ms1 )

matrix <- mdata$data
rownames(matrix) <- paste(mdata$row_info[,'protein_name'], mdata$row_info[,'id_number'])
colnames(matrix) <- colnames(mdata$data)

col_fun = circlize::colorRamp2(c(0, 1), c( "white", "orange"))

col_split = factor(paste(stringr::str_split(colnames(matrix), "_", n = 3, 
  simplify = TRUE)[,1]))

hm = ComplexHeatmap::Heatmap(matrix, 
  #column_split = col_split, 
  #column_title_gp = grid::gpar(fontsize = 8),
  column_gap = grid::unit(0.25, "cm"),
  column_order = levels(col_split),
  show_column_names = TRUE,
  #column_names_rot = 90,
  row_order = rownames(matrix), 
  show_row_names = TRUE,
  row_names_side = "left",
  rect_gp = grid::gpar(col = "white", lwd = 2 ), 
  col = col_fun 
  )
hm
```


## pmk1

```{r}
selected_pathway = pathways %>% filter(pathway == 'pmk1')

selected_ms1 = m[grepl(paste(selected_pathway$id_number, collapse="|"), ms1_tb$id),]
row_infos = data.frame(m_rownames = rownames(selected_ms1))
row_infos = row_infos %>% mutate(
  id_number = stringr::str_sub(m_rownames, end = 9)
  )
row_infos = left_join(row_infos, selected_pathway, by = "id_number", keep = FALSE)

mdata = list(row_info = row_infos, data = selected_ms1 )

matrix <- mdata$data
rownames(matrix) <- paste(mdata$row_info[,'protein_name'], mdata$row_info[,'id_number'])
colnames(matrix) <- colnames(mdata$data)

col_fun = circlize::colorRamp2(c(0, 1), c( "white", "salmon"))

col_split = factor(paste(stringr::str_split(colnames(matrix), "_", n = 3, 
  simplify = TRUE)[,1]))

hm1 = ComplexHeatmap::Heatmap(matrix, 
  #column_split = col_split, 
  #column_title_gp = grid::gpar(fontsize = 8),
  column_gap = grid::unit(0.25, "cm"),
  column_order = levels(col_split),
  show_column_names = TRUE,
  #column_names_rot = 90,
  row_order = rownames(matrix), 
  show_row_names = TRUE,
  row_names_side = "left",
  rect_gp = grid::gpar(col = "white", lwd = 2 ), 
  col = col_fun 
  )
hm1
```

## Sln1/HOG

```{r}
selected_pathway = pathways %>% filter(pathway == 'Sln1/HOG')

selected_ms1 = m[grepl(paste(selected_pathway$id_number, collapse="|"), ms1_tb$id),]
row_infos = data.frame(m_rownames = rownames(selected_ms1))
row_infos = row_infos %>% mutate(
  id_number = stringr::str_sub(m_rownames, end = 9)
  )
row_infos = left_join(row_infos, selected_pathway, by = "id_number", keep = FALSE)

mdata = list(row_info = row_infos, data = selected_ms1 )

matrix <- mdata$data
rownames(matrix) <- paste(mdata$row_info[,'protein_name'], mdata$row_info[,'id_number'])
colnames(matrix) <- colnames(mdata$data)

col_fun = circlize::colorRamp2(c(0, 1), c( "white", "darkolivegreen3"))

col_split = factor(paste(stringr::str_split(colnames(matrix), "_", n = 3, 
  simplify = TRUE)[,1]))

hm2 = ComplexHeatmap::Heatmap(matrix, 
  #column_split = col_split, 
  #column_title_gp = grid::gpar(fontsize = 8),
  column_gap = grid::unit(0.25, "cm"),
  column_order = levels(col_split),
  show_column_names = TRUE,
  #column_names_rot = 90,
  row_order = rownames(matrix), 
  show_row_names = TRUE,
  row_names_side = "left",
  rect_gp = grid::gpar(col = "white", lwd = 2 ), 
  col = col_fun 
  )
hm2
```


```{r}
ComplexHeatmap::draw(hm %v% hm1 %v% hm2)
```
## loop



## Sln1/HOG

```{r}

hm_list = list()

colours = c(
  'gold', 'orange', 'darkolivegreen3', 'deepskyblue4', 'darkgrey', 'salmon'
)
names(colours) = unique(pathways$pathway)

for (p in unique(pathways$pathway)) {
  #print(p)
  selected_pathway = pathways %>% filter(pathway == p)
  
  selected_ms1 = m[grepl(paste(selected_pathway$id_number, collapse="|"), ms1_tb$id),]
  row_infos = data.frame(m_rownames = rownames(selected_ms1))
  row_infos = row_infos %>% mutate(
    id_number = stringr::str_sub(m_rownames, end = 9)
    )
  row_infos = left_join(row_infos, selected_pathway, by = "id_number", keep = FALSE)
  
  mdata = list(row_info = row_infos, data = selected_ms1 )
  
  matrix <- mdata$data
  rownames(matrix) <- paste(mdata$row_info[,'protein_name'], mdata$row_info[,'id_number'])
  colnames(matrix) <- colnames(mdata$data)
  
  col_fun = circlize::colorRamp2(c(0, 1), c( "white", colours[p]))
  
  col_split = factor(paste(stringr::str_split(colnames(matrix), "_", n = 3, 
    simplify = TRUE)[,1]))
  
  hm = ComplexHeatmap::Heatmap(matrix, 
    #column_split = col_split, 
    #column_title_gp = grid::gpar(fontsize = 8),
    column_gap = grid::unit(0.25, "cm"),
    column_order = levels(col_split),
    show_column_names = TRUE,
    #column_names_rot = 90,
    row_order = rownames(matrix), 
    show_row_names = TRUE,
    row_names_side = "left",
    rect_gp = grid::gpar(col = "white", lwd = 2 ), 
    col = col_fun 
  )
  hm
  
  
  
  hm_list[[p]] = hm
}


```

```{r}
ComplexHeatmap::draw(hm_list$CpkA %v% hm_list$pmk1 %v% hm_list$`Sln1/HOG`)
```

